Tail associations in ecological variables and their impact on extinction risk

### 2020-01-21 MainText.Rmd
minr changes on read through

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable made in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic and complex, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms outlined above. 
By so doing we hope to open these potentially practically important questions to 
wider examination by other researchers, using a variety of models and perspectives. 
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula),
and our essential goal is to advance the idea that tail associations may
also be important in this context. 
Our questions are also an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. The variance, $\var$,
here and throughout, is the sample variance, with Bessel's correction (denominator $T-1$).
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. Each band is bounded by
the lines $x+y=2l_b$ and $x+y=2u_b$ described in Methods (for values of $l_b$ and $u_b$
that depend on the band to be used). This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models. These models were all described by @cohen1995unexpected. 
The models are also listed in Table \ref{tab_model_summary}, and 
are analyzed in detail in Appendix \ref{SM-model_stability}, where 
references for the models are also provided.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
at time step $25$.
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the Introduction, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_{min},r_{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,..0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for two
fixed values of $r$, one producing under- and one overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Differences were small, corresponding 
only to a few percentage points of additional extinction risk after 25 years. A 
potential reason for this outcome is elaborated in the Discussion.

@@

<!--future ideas start here-->
Our initial exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with $r_2$, whereas lower 
values of $r_1$ may not be strongly
associated with $r_2$ because $r_1$ trades off against both 
$r_2$ and $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, in
mean, variance, and possibly other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change, which is instead a change 
in marginal distributions. Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods.

Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas 
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
we used are closely related to copulas, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be useful for many applications. But to understand potential 
changes in the association 
of $X_i(t)$ and $X_j(t)$, unconfounded by changes in their marginal distributions,
it is necessary to use rank-based or copula methods [@Ghosh_copula,
@Genest2007].

>>>>

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable made in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic and complex, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms outlined above. 
By so doing we hope to open these potentially practically important questions to 
wider examination by other researchers, using a variety of models and perspectives. 
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula).
Our essential goal is to advance the idea that tail associations may
also be important in this context. 
Our questions are also an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$, divided by $T+1$, and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$, divided by $T+1$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. The variance, $\var$,
here and throughout, is the sample variance, with Bessel's correction (denominator $T-1$).
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the normalized ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. Each band is bounded by
the lines $x+y=2l_b$ and $x+y=2u_b$ described in Methods (for values of $l_b$ and $u_b$
that depend on the band to be used). This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2020).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

<!--density-dependence metapopulation model-->
To investigate Q2 from the Introduction, we considered spatialized, stochastic versions
of six density dependent population models, the Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, 
and Austin-Brewer models. These models were all described by @cohen1995unexpected. 
The models are also listed in Table \ref{tab_model_summary}, and 
are analyzed in detail in Appendix \ref{SM-model_stability}, where 
original references for the models are also provided.
All models took the general form $\vec{P} (t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{\text{th}}$ component $P_i(t)$ representing
the population density in the $i^{\text{th}}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{\text{th}}$ diagonal entry 
$\lambda_i(t)$ depending on the model used (Table \ref{tab_model_summary}). The matrix $D$
was an $n \times n$ dispersal matrix. For simplicity, 
we assumed the $n$ habitat patches were arranged, evenly spaced, in a line.
We considered both "local" and "global" dispersal among patches in different simulations. 
For local dispersal, a fraction $d$ of 
each population dispersed during each time step, equally distributed to
the two or one nearest neighbors. For global dispersal the same fraction $d$ dispersed
equally to the other $n-1$ patches. When $d$ was $0$, dispersal did not occur and 
rescue effects were not possible.

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
at time step $25$.
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution. Though these confidence intervals have limited
utility because they can be made arbitrarily narrow by increasing the number of simulations,
they are nonetheless presented to illustrate the confidence achieved with the number of 
simulations we performed.

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of the Introduction, 
we estimated extinction risk for each metapopulation model, using each form of tail 
association for the environmental noise,
using values of $r$ that included under- and overcompensatory dynamics,
as indicated above. We chose values of $r$ within a range, the
range specified as follows. The second part of Q2 (about how density dependence mediates the
influence of asymmetric tail associations on extinction risk) was addressed by comparing
results for under- and overcompensatory model regimes.
Defining $r_{min}$ as the minimum value of $r$ 
for which $P_e$ was non-negative; and defining
$r_{bf}$ as the value of $r$ at which a bifurcation occurs, the range of
$r$ used spanned 95\% of the interval ($r_{min},r_{bf}$). For details, see Appendix 
\ref{SM-model_stability}. The range of $r$ encompassed $r_c$, so included 
a transition from under- to overcompensatory dynamics. For each value of $r$
used within the range, we calculated extinction risk as indicated above.
Dispersal parameter values $d= 0,0.1,0.2,\ldots,0.5$ were also used in different 
simulations. We plotted extinction risk as a function of $d$ for several
fixed values of $r$, some producing under- and some overcompensatory dynamics.
For $d=0$ we also plotted extinction risk as a function of $r$. All code for 
this project is available at https://github.com/sghosh89/ERC.

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to the density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Differences were small, corresponding 
only to a few percentage points of additional extinction risk after 25 years. A 
potential reason for this outcome is elaborated in the Discussion.

@@

<!--future ideas start here-->
Our exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with $r_2$, whereas lower 
values of $r_1$ may not be strongly
associated with $r_2$ because $r_1$ trades off against both 
$r_2$ and $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

@@

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, in
mean, variance, and possibly other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change, which is instead a change 
in marginal distributions. Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based correlation methods.

Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas 
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
we used are closely related to copulas, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be useful for many applications. But to understand potential 
changes in the association 
of $X_i(t)$ and $X_j(t)$, unconfounded by changes in their marginal distributions,
it is necessary to use rank-based or copula methods [@Ghosh_copula;
@Genest2007].
